---
title: "Racial Force Probe"
author:
  - Monica Sainani
  - Kaavya Kalani
  - Aamishi Avarsekar
thanks: "Code and data are available at: https://github.com/kaavyakalani26/prison_health.git and replication aspects are available at: https://www.socialsciencereproduction.org/reproductions/82b709e2-4aa0-40e8-8abe-195c51ae739b/index"
date: today
date-format: long
abstract: "This research endeavors to scrutinize the well-being of individuals within the prison system, focusing on the prevalence of physical diseases, mental health issues, and mortality rates among incarcerated populations. The prison environment presents unique challenges that can significantly impact the overall health of individuals, making it imperative to investigate the multifaceted dimensions of prisoners' well-being. Utilizing data collected over the past [insert number of years], this study employs a comprehensive analysis to discern patterns, trends, and potential variations in the health outcomes of prisoners. By examining the intersection of physical and mental health alongside mortality rates, the research aims to contribute valuable insights into the current state of prisoners' well-being. This analysis not only serves to inform our understanding of the challenges faced by incarcerated individuals but also provides a foundation for evidence-based policy recommendations aimed at fostering improved conditions within correctional facilities. Through this exploration, we strive to contribute to the ongoing dialogue surrounding the intersection of criminal justice and public health, emphasizing the importance of compassionate and effective strategies for promoting the well-being of those within the prison system."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

#### Preamble ####
# Purpose: Read the cleaned dataset to use it for the analysis
# Author: Kaavya Kalani, Monica Sainani, Aamishi Avarsekar
# Date: 14 February 2024
# Contact: kaavya.kalani@mail.utoronto.ca, monica.sainani@mail.utoronto.ca, aamishi.avarsekar@mail.utoronto.ca
# License: MIT
# Pre-requisites: run scripts/01-data_cleaning.R

#### Workspace setup ####
# install.packages("tidyverse")
# install.packages("knitr")
# install.packages("here")
# install.packages("ggplot2")
# install.packages("janitor")
# install.packages("zoo")

library(ggplot2)
library(tidyverse)
library(knitr)
library(here)
library(janitor)
library(zoo)
```


# Introduction

Health and Safety of Prisoners Today

Individuals' health and safety in jail are critical aspects of the criminal justice system
that must be thoroughly examined. As we negotiate the complicated world of prisons in contemporary America,
it becomes clear that the physical and emotional health of jailed people is a major problem. This study aims to give a
complete analysis of the health difficulties encountered by inmates today, diving into the areas of physical health,
emotional well-being, death rates, and the subtle distinctions between male and female prisoners.

Physical Health of Prisoners: Unveiling a Disturbing Reality

The first aspect of our investigation is a thorough examination of the physical health of convicts.
Recent discoveries in America portray a troubling picture, demonstrating that a sizable proportion of offenders enter
correctional facilities with a variety of pre-existing physical health issues.
The jail environment, which is marked by overcrowding, restricted access to healthcare, and a lack of preventive
measures, frequently exacerbates these problems(Stevenson, B. (2021)). According to studies, the prison population has a high prevalence of
chronic diseases such as diabetes, hypertension, and infectious diseases, emphasizing the critical need to focus on
the physical well-being of jailed people.

The sad fact is that convicts have far lower health outcomes than the general population.
Access to medical care in prisons remains a recurrent issue, resulting in delayed diagnoses and insufficient treatment.
The cumulative effect of these conditions adds to a cycle of declining physical health among inmates, raising not just
ethical problems but also public health dangers when they are eventually released into society(M. Maruschak, L., & Berzofsky, M. (2015)).

Mental Health: The Silent Crisis Behind Bars

Moving beyond physical health, the second half of our investigation focuses on the sometimes-ignored topic of mental
well-being in the prison system. Mental health difficulties among inmates have reached crisis proportions, with a slew
of obstacles compounding the psychological toll of incarceration.(Edgemon, T. G., & Clay-Warner, J. (2019)). Suicide rates among convicts have become a major
problem, underlining the critical need for mental health services in correctional facilities.

Substance addiction exacerbates the mental health problem since many inmates take drugs to cope with the stresses of
confinement. The absence of comprehensive addiction treatment services in jails reinforces the cycle of substance misuse,
exacerbating mental health issues and impeding successful recovery attempts.(Edgemon, T. G., & Clay-Warner, J. (2019))

The mental health crisis among inmates emphasizes the significance of treating the underlying causes of psychological
anguish in correctional facilities. Initiatives to provide mental health support, provide counselling services, and
establish drug addiction treatment programs are critical to interrupting the cycle of mental health worsening among inmates.


From Suicides to Mortality Rates: Connecting the Dots

The third feature of our investigation focuses on the relationship between suicides and prison death rates.
Suicides in correctional facilities are not isolated instances; they are inextricably linked to larger concerns such as
mental health, poor healthcare, and the harsh reality of incarceration. Beyond suicides, inmates are always at risk of
death from a variety of causes, including murders, infectious infections, and other health difficulties.(Emily Widra, (2017))

Understanding the interconnectedness of these mortality variables is critical for designing holistic measures to improve
prisoner safety and well-being((Emily Widra, (2017)). It demands a comprehensive strategy that tackles both immediate concerns, such as
suicide prevention, as well as the larger structural factors that contribute to higher prison death rates.

Gender Disparities in Prison Health: Unveiling Unique Challenges

In the third portion of our investigation, we examine the subtle disparities in the health experiences of male and female
convicts. While many health issues are common, gender-specific characteristics play an important role in defining the health
outcomes of jailed people. Women inmates, for example, frequently confront particular reproductive health challenges,
as well as an increased risk of sexual assault while in correctional facilities.

Understanding and treating gender differences in prison health is critical for establishing focused therapies that
meet the unique requirements of both male and female inmates. Recognizing the unique issues experienced by each gender
group is an important step toward establishing a more equal and compassionate approach to healthcare in the prison system.

Conclusion: A Call for Holistic Reforms

To summarize, the health and safety of inmates today require immediate attention
and thorough improvements. The complex network of physical health issues, mental health crises,
mortality risks, and gender imbalances inside the jail system needs a comprehensive strategy.
Addressing these concerns necessitates not just urgent actions, such as increased healthcare access and mental health
assistance, but also a larger commitment to rethinking the purpose and consequences of incarceration on individual
well-being(Ghazanfari, H., & Farokhzadian, J. (2022)).This article intends to add to the continuing discussion about prison health by arguing for policies and
practices that prioritize the dignity, safety, and general well-being of people housed in correctional institutions.

# Data

# Results

```{r}
#| message: false
#| echo: false
#| label: fig-overall-death
#| fig-cap: Causes of deaths among the prisoners
```

```{r}
#| message: false
#| echo: false
#| label: fig-covid
#| fig-cap: Covid trends in the jail

pop <- c(1217352, 412783, 328239523)  # population denominators

# Read CSV file with the correct date format
x <- read.csv("data/raw_data/covidts01.csv", header = TRUE, stringsAsFactors = FALSE)
x$date <- as.Date(x$date, format = "%m/%d/%Y")  # Specify the correct date format

# Calculate rolling averages
pp <- 1e5 * zoo::rollapply((x$pp_cases[-1] - x$pp_cases[-132]), width = 7, FUN = mean) / pop[1]
sp <- 1e5 * zoo::rollapply((x$sp_cases[-1] - x$sp_cases[-132]), width = 7, FUN = mean) / pop[2]
gp <- 1e5 * zoo::rollapply((x$gp_cases[-1] - x$gp_cases[-132]), width = 7, FUN = mean) / pop[3]

# Combine data into a data frame
ts_data <- data.frame(
  date = rep(x$date[-(1:7)], 3),
  new_cases_per_100k = c(pp, sp, gp),
  group = rep(c("Incarcerated", "Prison Staff", "General Population"), each = length(pp))
)

# Create the figure using ggplot2 with solid lines
ggplot(ts_data, aes(x = date, y = new_cases_per_100k, color = group)) +
  geom_line(size = 0.8) +  # Adjust the line thickness here (e.g., size = 0.8)
  scale_x_date(date_labels = "%B", date_breaks = "1 month", limits = c(min(ts_data$date), max(ts_data$date))) +
  scale_y_continuous(limits = c(0, 150), breaks = seq(0, 150, 25)) +
  labs(x = NULL, y = "New cases per 100,000", title = "Seven-Day Moving Average of New Daily COVID-19 Cases per 100,000 Among
Those in Prison, Prison Staff, and in the General Population: May to September
2020") +
  theme_minimal() +
  geom_vline(xintercept = as.numeric(as.Date(c("2020-04-01", "2020-05-01", "2020-06-01", "2020-07-01", "2020-08-01", "2020-09-01"))),
             linetype = "dashed", color = "gray") +
  annotate("text", x = as.Date(c("2020-04-01", "2020-05-01", "2020-06-01", "2020-07-01", "2020-08-01", "2020-09-01")),
           y = 150, label = c("April 2020", "May 2020", "June 2020", "July 2020", "August 2020", "September 2020"),
           vjust = -0.5, hjust = 0, color = "gray") +
  scale_color_manual(values = c("blue", "red", "green"))

```

```{r}
#| message: false
#| echo: false
#| label: fig-death-vs-gender
#| fig-cap: Causes of deaths among the prisoners by gender

health_data_count_by_demographics <- read_csv(file = here("data/raw_data/msfp0116stt09data.csv"))

health_demographic_data_clean <-
  clean_names(health_data_count_by_demographics) |>
  select("x1", "cancer", "heart_disease", "liver_disease",
         "respiratory_disease", "aids_related", "all_other_illnesses", "suicide",
         "drug_alcohol_intoxication", "accident", "homicide_d")

# Create a mapping for all causes
cause_mapping <- c(
  "cancer" = "Cancer",
  "heart_disease" = "Heart Disease",
  "liver_disease" = "Liver Disease",
  "respiratory_disease" = "Respiratory Disease",
  "aids_related" = "Aid related",
  "all_other_illnesses" = "All other illnesses",
  "suicide" = "Suicide",
  "drug_alcohol_intoxication" = "Drug/Alcohol Intoxication",
  "accident" = "Accident",
  "homicide_d" = "Homicide"
)

# Rename the causes in the dataset
health_demographic_data_clean <- health_demographic_data_clean %>%
  rename_with(~ cause_mapping[.], -x1)

gender_health_data <- health_demographic_data_clean %>% filter(x1 %in% c("Male", "Female"))

gender_health_data_long <- pivot_longer(gender_health_data, cols = -x1, names_to = "Cause", values_to = "Count")

# Calculate proportions within each category
gender_health_data_long <- gender_health_data_long %>%
  group_by(x1) %>%
  mutate(Proportion = Count / sum(Count))

# Plot the proportions graph using geom_bar with position = "fill"
ggplot(gender_health_data_long, aes(x = x1, y = Proportion, fill = Cause)) +
  geom_bar(stat = "identity", position = "dodge", color = "black", size = 0.5) +  # Add black borders
  labs(title = "Proportions of Causes of Death by Category",
       x = "Gender",
       y = "Proportion") +
  scale_fill_discrete(labels = cause_mapping) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#| message: false
#| echo: false
#| label: fig-death-vs-race
#| fig-cap: Causes of deaths among the prisoners by race

# Read data
health_data_count_by_demographics <- read_csv(file = here("data/raw_data/msfp0116stt09data.csv"))

# Clean and organize data
health_demographic_data_clean <-
  clean_names(health_data_count_by_demographics) |>
  select("x1", "cancer", "heart_disease", "liver_disease",
         "respiratory_disease", "aids_related", "all_other_illnesses", "suicide",
         "drug_alcohol_intoxication", "accident", "homicide_d")

# Pivot data to long format
health_demographic_data_long <- pivot_longer(health_demographic_data_clean, cols = -x1, names_to = "Cause", values_to = "Count")

# Filter data for specific demographics (e.g., "Male" and "Female")
filtered_df <- health_demographic_data_clean[health_demographic_data_clean$x1 %in% c("Male", "Female"),]

# Filter data for specific races (e.g., "White", "Black", "Hispanic", "Other")
race_health_data <- health_demographic_data_clean %>% filter(x1 %in% c("White", "Black", "Hispanic", "Other"))

# Pivot data to long format for race
race_health_data_long <- pivot_longer(race_health_data, cols = -x1, names_to = "Cause", values_to = "Count")

# Calculate proportions instead of counts
race_health_data_long <- race_health_data_long %>%
  group_by(x1) %>%
  mutate(Proportion = Count / sum(Count))

# Create a mapping for all causes
cause_mapping <- c(
  "cancer" = "Cancer",
  "heart_disease" = "Heart Disease",
  "liver_disease" = "Liver Disease",
  "respiratory_disease" = "Respiratory Disease",
  "aids_related" = "Aid related",
  "all_other_illnesses" = "All other illnesses",
  "suicide" = "Suicide",
  "drug_alcohol_intoxication" = "Drug/Alcohol Intoxication",
  "accident" = "Accident",
  "homicide_d" = "Homicide"
)

# Rename the causes in the dataset
race_health_data_long <- race_health_data_long %>%
  mutate(Cause = cause_mapping[as.character(Cause)])

# Create a bar plot showing proportions with black borders
ggplot(race_health_data_long, aes(x = x1, y = Proportion, fill = Cause)) +
  geom_bar(stat = "identity", position = "dodge", color = "black", size = 0.5) +  # Add black borders
  labs(title = "Proportions of Causes of Death by Category",
       x = "Category",
       y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

# Discussion

## Weaknesses and next steps

# Appendix

# References
